The impact of early massive mergers on the chemical evolution of Milky Way-like galaxies: insights from NIHAO-UHD simulations

### 2023-04-04 src/tex/ms.tex
[showyourwork] overleaf sync

\def \etal {et~al.~}
\newcommand{\MKN}[1]{{\color{blue}{\bf MKN}:~ #1}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[Early massive accretion events in MW-mass galaxies]{The influence of Gaia-Sausage-Enceladus-like merger events on the chemical evolution of galaxy discs}
%
%{The influence of Gaia-Sausage-Enceladus-like merger events on the chemo-dynamics of galaxy discs}
%The influence of massive early merger events on the chemo-dynamics of galaxy discs

\author[T. Buck] {Tobias Buck$^{1,2}$\thanks{E-mail: tobias.buck@iwr-uni-heidelberg.de.de}, Aura Obreja$^{3}$, \etal \\  %Andrea V. Macci\`o$^{4,5}$, \etal \\
%, Aaron A. Dutton$^{2}$, Andrea V. Macci\`o$^{2,1}$\\
%
$^1$Universit\"at Heidelberg, Interdisziplin\"ares Zentrum f\"ur Wissenschaftliches Rechnen, Im Neuenheimer Feld 205, \\ D-69120 Heidelberg, Germany\\
$^2$Universit\"at Heidelberg, Zentrum f\"ur Astronomie, Institut f\"ur Theoretische Astrophysik, Albert-Ueberle-Straße 2, \\ D-69120 Heidelberg, Germany\\
$^3$Universit\"ats-Sternwarte M\"unchen, Scheinerstraße 1, D-81679 M\"unchen, Germany%\\
%$^4$New York University Abu Dhabi, PO Box 129188, Saadiyat Island, Abu Dhabi, United Arab Emirates\\
%$^5$Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, 69117 Heidelberg, Germany\\
}

@@

\begin{abstract}
awesome science! \\
keep on reading \\
and cite me, please!\\
we have really great results!\\
the greatest results you will see today...\\
I hope this is enough to fill some lines \\
and add some space that might be filled by an actual abstract.\\
Let's see \\

\end{abstract}

@@

Our Milky Way (MW) Galaxy is the best-studied galaxy in the
universe and provides the most stringent constraints of galaxy formation
models \citep[e.g.,][]{Guedes2011, Wetzel2016, Grand2017, Buck2020a}. Therefore, a thorough understanding of the components of the Galaxy, their dynamical evolution and formation channel is of great interest.

The Galaxy's main body, the rotationally-supported stellar disk harbors multiple components or populations when dissected in the space of chemistry, kinematics, spatial extent, and age \citep[e.g.,][]{Gilmore1983, Norris1985, Chiba2000, Nissen2010, Bovy2012, Haywood2013}; see \citet{Rix2013} and \citet{Bland-Hawthorn2016} for recent reviews. %There exist a dichotomy in both spatial structure and stellar chemistry. Spatially, MW's disk can be separated into a thin and a thick disk with different scale heights. Chemically, a clear bimodality in the abundance of $\alpha$ elements relative to iron ([$\alpha$/Fe]) has been found. In general, the high$-\alpha$ population has older ages and a larger scale-height than the low$-\alpha$ population. However, the mapping between the high$-\alpha$ (low$-\alpha$) and the thick (thin) disks is not perfect.

Thanks to the great improvements in the quality and volume of astrometric datasets provided by the {\it Gaia} mission in combination with chemical abundances and radial velocities from large spectroscopic surveys, (e.g. \textsc{apogee}; \textsc{galah}; \textsc{h3}; \textsc{LAMOST}; \citealt{Majewski2017}, \citealt{Martell2017,Buder2021}, \citealt{Conroy2019}, \citealt{Zhao2012}, respectively) we are now able to identify and link coherent chemodynamical structures in the solar neighbourhood to ancient merger events several billion years back in MW's history (e.g. \citealt{Helmi2020} for a review). By now a large number of chemodynamical clusters associated either with distinct in-situ components at the time of birth of the MW's stellar disk \citep[e.g.][]{Belokurov2022,Conroy2022} or associated with distinct accreted systems have been identified in the high-dimensional space of orbital parameters and stellar abundances \citep[e.g.,][]{Myeong2019, Naidu2020, Horta2021, Buder2022}.

Most notably, there exists an excess of stars on radial orbits in the local stellar halo around the Sun which is often referred to as {\it Gaia}-Sausage-Enceladus (GSE; \citealt{Belokurov2018, Helmi2018}; see also \citealt{Nissen2010, Koppelman2018, Haywood2018}). This stellar overdensity has been identified as the observable signature of a massive ($M_{\star} \simeq 10^{9}$ $\rm{M_{\odot}}$) early ($8-11$ Gyr ago; e.g. \citealt{Vincenzo2019, Belokurov2020, Naidu2021, Xiang2022}) accretion event. GSE is by far the most significant merger in MW's history and contributed approximately two-thirds of MW's stellar halo stars on highly-eccentric orbits \citep[e.g.][]{Mackereth2020}. GSE stars inhabit a sausage-like distribution in the radial-azimuthal velocity distribution \citep{Brook2003, Belokurov2018} and appear to be more metal-poor and less $\alpha$-enhanced than the redder halo counterpart \citep{Haywood2018, Helmi2018}. The redder sequence of the Galactic halo is thought to be the result of proto-galactic disc stars being dynamically ejected into the halo during the GSE merger and has been associated with an event termed {\it the Splash} by \cite{Belokurov2020, Bonaca2020} which has been initially found by other earlier studies \citep{Bonaca2017, Haywood2018, DiMatteo2019, Gallart2019}.

The existence of the GSE merger and its strong impact on the structure of the proto-MW demands a thorough study of the processes that shape MW's chemo-dynamical evolution during such a merger in order to interpret the observational data at hand. In order to connect present day observables to Gyr old events we need tools like cosmological simulations that model galactic mass growth, mergers and subsequent star formation plus chemical enrichment self-consistently. Here, enough resolution to follow the internal disk dynamics and suppress spurious heating ($N \gtrsim 10^6$; e.g. \citealt{Sellwood2013,Ludlow2019,Ludlow2021}) as well as resolving the multi-phase, dense structure of the interstellar medium (ISM) combined with a proper model for chemical enrichment \citep[e.g.][]{Buck2021} is key to accurately capture their birth kinematics (e.g. \citealt{Ratcliffe2022, Lu2022}) and the subsequent dynamical evolution of the disk (e.g. \citealt{House2011, Bird2013, Buck2019b}).

\begin{figure*}
    \script{merger_ratios.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/merger_ratios.pdf}
        \vspace*{-1.75em}
        \caption{Gas mass (blue line, left axis) and gas merger ratio (orange line, right axis) as a function of time. We identify several gas rich mergers that contribute around 10\% of gas to the main galaxy at various times of 2-4 Gyr and a late time merger starting around 9.5 Gyr that contributes 10\% of gas and continues for several peri-center passages and goes on for $\sim2$ Gyr.
        }
        \label{fig:merger_ratio}
    \end{centering}
\end{figure*}

Thanks to recent advances in computing power and progress in numerical methods modern cosmological zoom simulations now meet the aforementioned requirements and are able to model individual MW-mass galaxies by sampling varying environments and formation scenarios (e.g. \citealt{Sawala2016, Grand2017, Buck2018, Font2020, Applebaum2021, Agertz2021, Bird2021, Khoperskov2022-InSitu}). This enables us to reconstruct our Galaxy's formation history by linking the occurrence of chemodynamical patterns at $z=0$ to specific events in the Galaxy's \citep[e.g.][]{Bignone2019, Fattahi2019, Grand2020, Elias2020, Dillamore2022, Khoperskov2022a, Rey2022}.

While GSE's strong impact on MW's stellar halo is by now well established, more recently studies have further investigated its defining influence on the formation of MW's thin and thick disk \citep[e.g.]{Grand2020,Ciuca2022,Orkney2022,Rey2023}. These studies suggest that the GSE merger strongly shaped the formation of MW's thick disk agreeing with thick disk formation scenarios from early simulation results by \cite{Brook2004, Brook2006}. The gas-rich merger has a two-fold effect on MW's stellar disk: (i) Firstly, it heats part of the existing proto-disc stars dynamically ejecting them into the halo and creating the Splash. (ii) Secondly, the merger provides fresh gas to the central galactic regions triggering a starburst that eventually forms the younger thick disc. After the merge, the thin disc then forms from the accretion of metal-poor gas in an inside-out, upside-down fashion \citep[e.g.,][]{Bird2013, Grand2018, Buck2020a}. \citet{Ciuca2021} qualitatively confirmed this picture using APOGEE DR14 data and selecting stellar populations around the solar radius.

Interestingly, when inferring the birth radii of MW disk stars, \citet{Lu2022b} found a steepening of the metallicity gradient at the time of the GSE merger using data from LAMOST combined with {\it Gaia} eDR3. Here we use the NIHAO-UHD suite \citep{Buck2020a} of cosmological hydrodynamical simulations of MW-mass galaxies in order to investigate the influence of early merger events on the chemo-dynamics of the stellar disk. By using 4 different simulations from the NIHAO-UHD suite we investigate independently sampled environments that follow 4 different formation scenarios. Especially we focus on the frequency of massive early mergers and their detailed impact on the formation of the stellar disk. This paper is structures as follows: In \S~2 we describe the simulations on which we base our analysis, followed by a presentation of the impact of early massive mergers on the metallicity gradient of the cold gas disk in \S~3. In \S~4 we turn to analyse where merger stars and gas end up in the age-metallicity relation  and the [$\alpha$/Fe] vs. [Fe/H] plane of the main galaxy. In \S~5 we summarize and conclude our findings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods} \label{sec:simulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For this work we use four simulations from the NIHAO-UHD suite \citep{Buck2020a} part of the Numerical Investigation of a Hundred Astronomical Objects (NIHAO) simulation suite \citep{Wang2015}. Parts of this simulation suite have previously been used to study the build-up of MW's peanut-shaped bulge \citep{Buck2018,Buck2019}, investigate the stellar bar properties \citep{Hilmi2020}, infer the MW's dark halo spin \citep{Obreja2022}, study the dwarf galaxy inventory of MW mass galaxies \citep{Buck2019a} or investigate the age-metallicity relation of MW disk stars \citep{Lu2022} including the chemical bimodality of disk stars \citep{Buck2020}, their abundances \citep{Lu2022a} and the origin of very metal-poor stars inside the stellar disk \citep{Sestito2021}.
Comparing the properties of these galaxies to observations of the MW and local disk galaxies from the SPARC data \citep{Lelli2016} \citet{Buck2020a} showed that simulated galaxy properties agree well with observations.

The simulations assume cosmological parameters from the \cite{Planck}, namely: \OmegaM= 0.3175, \OmegaL= 0.6825, \Omegab= 0.049, H${_0}$ = 67.1\kms\Mpc$^{-1}$, \sig8 = 0.8344. Initial conditions are created the same way as for the original NIHAO runs \citep[see][]{Wang2015} using a modified version of the \texttt{GRAFIC2} package \citep{Bertschinger2001,Penzo2014}. The mass resolution of these simulation ranges between $m_{\rm dark}\sim1.5 - 5.1\times10^5 \Msun$ for dark matter particles and $m_{\rm gas}\sim2.8 - 9.4\times10^4 \Msun$ for the gas particles. The corresponding force softenings are $\epsilon_{\rm dark}=414 - 620$ pc for the dark matter and $\epsilon_{\rm gas}=177 - 265$ pc for the gas and star particles. However, the adaptive smoothing length scheme implies that $h_{\rm  smooth}$ can be as small as $\sim30$ pc in the disk mid-plane. Stellar particles are born with an initial mass of $1/3\times m_{\rm{gas}}$ and are subject to massloss according to stellar evolution models as detailed in \citet{Stinson2013}. 
The simulation setup, star formation and feedback implementations are described in detail in the introductory paper \citep{Buck2020a} but for completeness we summarise them below.

@@

Following \citet{Stinson2013} two modes of stellar feedback are implemented: (i) the energy input from young massive stars, e.g. stellar winds and photo ionization, prior to any supernovae explosions, thus termed \textit{early stellar feedback (ESF)}. This mode consists of the total stellar luminosity ($2 \times 10^{50}$ erg of thermal energy per $M_{\odot}$) of the entire stellar population with an efficiency for coupling the energy input of $\epsilon_{\rm ESF}=13\%$ \citep{Wang2015}. (ii) supernova explosions implemented using the blastwave formalism as described in \citet{Stinson2006} and making use of a delayed cooling formalism for particles inside the blast region following \citet{McKee1977} in order to account for the adiabatic expansion of the supernova.
Finally, we adopted a metal diffusion algorithm between particles as described in \citet{Wadsley2008}.

\begin{figure*}
    \script{half_mass_radius_metal_gradient.py}
    \begin{centering}
     \includegraphics[width=\linewidth]{figures/half_mass_radius_metal_gradient.pdf}
        \vspace*{-1.75em}
        \caption{
            Half mass radius (orange line, right axis) and metallicity gradient (blue line, left axis) of the cold gas as a function of time. The sharp increases in the cold gas size around the times of gas rich mergers coincide with a sudden steepening of the metallicity gradient of $\sim0.02$ dex/kpc. The addition of relatively unenriched cold gas by the merging satellite leads to a sudden increase of the star-forming cold gas  mass in the outskirts of the galaxies while the central parts are unaffected. This steepens the gradient.
        }
        \label{fig:half_mass}
    \end{centering}
\end{figure*}

%\begin{figure}
%    \script{half_mass_radius_gas.py}
%    \begin{centering}
%        \includegraphics[width=\linewidth]{figures/279e12_gas_size.pdf}
%        \caption{
%            Half mass radius of the cold gas as a function of time. The sharp increases in the cold gas size around 2 Gyr and 9 Gyr indicate the influence of massive gas rich merger events on the galaxy. The addition of fresh cold gas leads to a sudden increase in cold gas half mass size.
%        }
%        \label{fig:half_mass}
%    \end{centering}
%\end{figure}

@@

We start our analysis on the influence of Gaia-Sausage-Enceladus-like merger
events on the chemo-dynamics of MW-like galaxy disks by looking at the merger history of the four galaxies. Figure~\ref{fig:merger_ratio} shows the cumulative gas mass growth (blue line, left axis) accompanied by the gaseous merger ratio (orange line, right axis) defined as the ratio of gas mass of the merging satellite over the gas mass of the main galaxy. We see that after an initial phase ($0-\sim2.5$ Gyr) of rapid gas mass growth accompanied by violent, relatively gas rich mergers there are several other mergers at later times that contribute more than 10\% in gas leading to sudden jumps in the gas mass of the main galaxy (blue line). 
Especially for g2.79e12 (left most panel) and g7.08e11 (right most panel) we see some late time ($\sim10$ Gyr) gas rich merger while the other two galaxies do not show any gas rich merger after $5$ Gyr for g8.26e11 and $7$ Gyr for g7.55e11.

\subsection{The impact of gas rich merger on the metallicity gradient} \label{sec:merger}

The sudden increase of gas mass due to these gas rich merger leads to a sudden increase in the half mass radius ($\gsim20$ kpc) of the cold gas disk of the main galaxy as the orange line (right axis) in Figure~\ref{fig:half_mass} shows. Simultaneously we plot the metallicity gradient of the cold gas disk of the main galaxy as a function of time (blue line in Fig.~\ref{fig:half_mass}, left axis) as measured by fitting a straight line to the radial metallicity profile omitting the central $2.5$ kpc. In order to minimize short time scale fluctuations we plot the running average with a window size of 10 corresponding to a time window of $\sim600$Myr.

>>>>

\def \etal {et~al.~}
\newcommand{\MKN}[1]{{\color{blue}{\bf MKN}:~ #1}}
\newcommand{\T}[1]{{\color{blue}{\bf T}:~ #1}}
\newcommand{\LL}[1]{\textbf{\textcolor{red}{LL: #1}}}
\newcommand{\br}[1]{\textbf{\textcolor{magenta}{BR: #1}}}
\newcommand{\AO}[1]{\textbf{\textcolor{green}{AO: #1}}}
\newcommand{\AVM}[1]{\textbf{\textcolor{cyan}{AVM: #1}}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[Early massive accretion events in MW-mass galaxies]{The impact of early massive mergers on the chemical evolution of Milky Way-like galaxies: insights from NIHAO-UHD simulations}
%
%{The influence of Gaia-Sausage-Enceladus-like merger events on the chemo-dynamics of galaxy discs}
%The influence of massive early merger events on the chemo-dynamics of galaxy discs

\author[T. Buck] {Tobias Buck$^{1,2}$\thanks{E-mail: tobias.buck@iwr-uni-heidelberg.de.de}, Aura Obreja$^{3}$, Bridget Ratcliffe$^{4}$, Yuxi(Lucy) Lu$^{5, 6}$, 
\newauthor Ivan Minchev$^{4}$, Andrea V. Macci\`o$^{7,8}$ \\  %, \etal \\
%, Aaron A. Dutton$^{2}$, Andrea V. Macci\`o$^{2,1}$\\
%
$^1$Universit\"at Heidelberg, Interdisziplin\"ares Zentrum f\"ur Wissenschaftliches Rechnen, Im Neuenheimer Feld 205, \\ D-69120 Heidelberg, Germany\\
$^2$Universit\"at Heidelberg, Zentrum f\"ur Astronomie, Institut f\"ur Theoretische Astrophysik, Albert-Ueberle-Straße 2, \\ D-69120 Heidelberg, Germany\\
$^3$Universit\"ats-Sternwarte M\"unchen, Scheinerstraße 1, D-81679 M\"unchen, Germany\\
${4}$Leibniz Institute for Astrophysics Potsdam, An der Sternwarte 16, 14482 Potsdam, Germany\\
$^{5}$Department of Astronomy, Columbia University, 550 West 120$^{\rm th}$ Street, New York, NY, USA\\
$^{6}$American Museum of Natural History, Central Park West, Manhattan, NY, USA\\
$^7$New York University Abu Dhabi, PO Box 129188, Saadiyat Island, Abu Dhabi, United Arab Emirates\\
$^8$Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, 69117 Heidelberg, Germany
}

@@

\begin{abstract}
Recent observations of the Milky Way (MW) found an unexpected steepening of the star-forming gas metallicity gradient around the time of the Gaia-Sausage-Enceladus (GSE) merger event.
Here we investigate the influence of early ($t_{\rm{merger}}\lesssim5$ Gyr) massive ($M_{\rm{gas}}^{\rm{merger}}/M_{\rm{gas}}^{\rm{main}}(t_{\rm{merger}})\gsim10\%$) merger events such as the Gaia-Sausage Enceladus merger in the MW on the evolution of the cold gas metallicity gradient. We use the NIHAO-UHD suite of cosmological hydrodynamical simulations of MW-mass galaxies to study the frequency of massive early mergers and their detailed impact on the morphology and chemistry of the gaseous disks. 
We find a strong steepening of the metallicity gradient at early times for all four galaxies in our sample which is caused by a sudden increase in the cold gas disk size (up to a factor of 2) in combination with the supply of un-enriched gas ($\sim0.75$ dex lower compared to the main galaxy) by the merging dwarf galaxies.
The mergers mostly affect the galaxy outskirts and lead to an increase in cold gas surface density of up to 200\% outside of $\sim8$ kpc. The addition of un-enriched gas breaks the self-similar enrichment of the inter-stellar-medium and causes a dilution of the cold gas in the outskirts of the galaxies. 
The accreted stars and the ones formed later out of the accreted gas inhabit distinct tracks offset to lower [$\alpha$/Fe] and [Fe/H] values compared to the main galaxy's stars. We find that such mergers can contribute significantly to the formation of a second, low-$\alpha$ sequence as is observed in the MW.
\end{abstract}

@@

Our Milky Way (MW) Galaxy is the best-studied galaxy in the
universe and provides some of the most stringent constraints of galaxy formation
models \citep[e.g.,][]{Guedes2011, Wetzel2016, Grand2017, Hopkins2018, Buck2020a}. Therefore, a thorough understanding of the components of the Galaxy, their dynamical evolution and formation channel is of great interest.

The Galaxy's main body, the rotationally-supported stellar disk harbors multiple components or populations when dissected in the space of chemistry, kinematics, spatial extent, and age \citep[e.g.,][]{Gilmore1983, Norris1985, Chiba2000, Nissen2010, Bovy2012, Haywood2013}; see \citet{Rix2013} and \citet{Bland-Hawthorn2016} for recent reviews.

Thanks to the great improvements in the quality and volume of astrometric datasets provided by the {\it Gaia} mission \citep{Gaia} in combination with chemical abundances and radial velocities from large spectroscopic surveys, (e.g. \textsc{apogee}; \textsc{galah}; \textsc{h3}; \textsc{LAMOST}; \citealt{Majewski2017}, \citealt{Martell2017,Buder2021}, \citealt{Conroy2019}, \citealt{Zhao2012}, respectively) we are now able to identify and link coherent chemodynamical structures in the MW to ancient merger events several billion years back in its history (e.g. \citealt{Helmi2020} for a review). By now a large number of chemodynamical clusters associated either with distinct in-situ components at the time of birth of the MW's stellar disk \citep[e.g.][]{Belokurov2022,Conroy2022,Rix2022} or associated with distinct accreted systems have been identified in the high-dimensional space of orbital parameters and stellar abundances \citep[e.g.,][]{Myeong2019, Naidu2020, Horta2021, Buder2022}.

Most notably, there exists an excess of stars on radial orbits in the local stellar halo around the Sun which is often referred to as {\it Gaia}-Sausage-Enceladus (GSE; \citealt{Belokurov2018, Helmi2018}; see also \citealt{Nissen2010, Koppelman2018, Haywood2018}). This stellar overdensity has been identified as the observable signature of a massive ($M_{\star} \simeq 10^{9}$ $\rm{M_{\odot}}$) early ($8-11$ Gyr ago; e.g. \citealt{Vincenzo2019, Belokurov2020, Naidu2021, Xiang2022}) accretion event. GSE is by far believed to be the most significant merger in MW's history and contributed approximately two-thirds of MW's stellar halo stars on highly-eccentric orbits \citep[e.g.][]{Mackereth2020}. GSE stars inhabit a sausage-like distribution in the radial-azimuthal velocity distribution \citep{Brook2003, Belokurov2018} and appear to be more metal-poor and less $\alpha$-enhanced than the redder halo counterpart \citep{Haywood2018, Helmi2018}. The redder sequence of the Galactic halo is thought to be the result of proto-galactic disc stars being dynamically ejected into the halo during the GSE merger and has been associated with an event termed {\it the Splash} by \cite{Belokurov2020, Bonaca2020} which has been initially found by other earlier studies \citep{Bonaca2017, Haywood2018, DiMatteo2019, Gallart2019}.

The existence of the GSE merger and its strong impact on the structure of the proto-MW demands a thorough study of the processes that shape MW's chemo-dynamical evolution during such a merger in order to interpret the observational data at hand. In order to connect present day observables to Gyr old events we need tools like cosmological simulations that model galactic mass growth, mergers and subsequent star formation plus chemical enrichment self-consistently. Here, one prerequisite to follow the internal disk dynamics and suppress spurious heating as well as resolving the multi-phase, dense structure of the interstellar medium (ISM) is enough resolution ($N \gtrsim 10^6$; e.g. \citealt{Sellwood2013,Ludlow2019,Ludlow2021}). At the same time a proper model for chemical enrichment \citep[e.g.][]{Buck2021} is needed to accurately capture the star's birth kinematics (e.g. \citealt{Ratcliffe2022, Lu2022}) and the subsequent dynamical evolution of the disk (e.g. \citealt{House2011, Bird2013, Buck2019b}).

Thanks to recent advances in computing power and progress in numerical methods modern cosmological zoom simulations now meet the aforementioned requirements and are able to model individual MW-mass galaxies by sampling varying environments and formation scenarios (e.g. \citealt{Sawala2016, Grand2017, Buck2018, Font2020, Applebaum2021, Agertz2021, Bird2021, Khoperskov2022-InSitu}). This enables us to reconstruct our Galaxy's formation history by linking the occurrence of chemodynamical patterns at $z=0$ to specific events in the Galaxy's evolution \citep[e.g.][]{Bignone2019, Fattahi2019, Grand2020, Elias2020, Dillamore2022, Khoperskov2022a, Rey2022}.

While GSE's strong impact on the MW's stellar halo is well established, more recently studies have further investigated its defining influence on the formation of the MW's thin and thick disk \citep[e.g.][]{Grand2020,Ciuca2022,Orkney2022,Rey2023}. These studies suggest that the GSE merger strongly shaped the formation of the MW's thick disk, agreeing with thick disk formation scenarios from early simulation results by \cite{Brook2004, Brook2006}. The gas-rich merger has a two-fold effect on MW's stellar disk: (i) it heats part of the existing proto-disc stars, dynamically ejecting them into the halo and creating the Splash. (ii) The merger provides fresh gas to the central galactic regions, triggering a starburst \citep[potentially after first causing a pause in star formation;][]{Vincenzo2019} that eventually forms the younger thick disc. After the merger, the thin disc then forms from the accretion of metal-poor gas in an inside-out, upside-down fashion \citep[e.g.,][]{Bird2013, Minchev2013, Grand2018, Buck2020a}. Using stellar populations around the solar radius from the APOGEE DR14 data \citet{Ciuca2021} qualitatively confirmed this picture.

It has recently become possible to infer the birth radii, $R_{\rm{b}}$, of stars in the Milky Way in a model-free manner. \cite{Minchev2018} devised a simple new technique for obtaining $R_{\rm{b}}$ in any observational dataset with precise age and metallicity measurements. This method simultaneously recovered $R_{\rm{b}}$ and the Galactic disk metallicity evolution with radius and time, [Fe/H]$(R,\tau)$, by varying the metallicity slope as function of time so as to keep the resulting $R_{\rm{b}}$ distributions physically meaningful. More recently, \cite{Lu2022b} found that a linear relation exists between the metallicity scatter and the metallicity birth gradient evolution with cosmic time, thus making the $R_{\rm{b}}$ determination fully self-consistent. The results of the above work have opened up a new way of studying the chemo-dynamical evolution of the Milky Way, where instead of relying on forward modeling over cosmic time (cosmological simulation or semi-analytical modeling), one gets there directly from the data with few prior assumptions on the detailed chemical enrichment.

\begin{figure*}
    \script{merger_ratios.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/merger_ratios.pdf}
        \vspace*{-1.75em}
        \caption{Gas mass (blue line, left axis) and gas merger ratio (orange line, right axis) as a function of cosmic time. The gray shaded areas mark the times when mergers heavily impact the gas disk size (see also Fig.~\ref{fig:half_mass}). Most of these mergers are at early times (2-4 Gyr), with the exceptions of g7.08e11 and g2.79e12, which also have a late time merger each.}
        \label{fig:merger_ratio}
    \end{centering}
\end{figure*}

Interestingly, when inferring the birth radii of MW disk stars, \citet{Lu2022b} found a steepening of the metallicity gradient at the time of the GSE merger using data from LAMOST DR7 combined with {\it Gaia} eDR3. When evaluating the chemical evolution across birth radii and lookback time using APOGEE DR17 red giant branch disk stars, Ratcliffe et al. submitted reported a similar steepening in the metallicity gradient at $\sim9$ Gyr ago, with a similar trend also appearing in other [X/H] abundances. Here we use the NIHAO-UHD suite of cosmological hydrodynamical simulations of MW-mass galaxies \citep{Buck2020a} %in order 
to investigate the influence of early merger events on the chemo-dynamics of the stellar disk. By using 4 different simulations from the NIHAO-UHD suite we investigate independently sampled environments that follow 4 different formation scenarios. We focus especially on the frequency of massive early mergers and their detailed impact on the formation of the stellar disk. This paper is structured as follows: In \S~2 we describe the simulations on which we base our analysis, followed by a presentation of the impact of early massive mergers on the metallicity gradient of the cold gas disk in \S~3. In \S~4 we turn to analyse where merger stars and gas end up in the age-metallicity relation  and the [$\alpha$/Fe] vs. [Fe/H] plane of the main galaxy. In \S~5 and \S~6 we summarize and conclude our findings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Simulations} \label{sec:simulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For this work we use four simulations from the NIHAO-UHD suite \citep{Buck2020a} part of the Numerical Investigation of a Hundred Astronomical Objects (NIHAO) simulation suite \citep{Wang2015}. Parts of this simulation suite have previously been used to study the build-up of MW's peanut-shaped bulge \citep{Buck2018,Buck2019}, investigate the stellar bar properties \citep{Hilmi2020}, infer the MW's dark halo spin \citep{Obreja2022}, study the dwarf galaxy inventory of MW mass galaxies \citep{Buck2019a} or investigate the age-metallicity relation of MW disk stars \citep{Lu2022} including the chemical bimodality of disk stars \citep{Buck2020}, their abundances \citep{Lu2022a} and the origin of very metal-poor stars inside the stellar disk \citep{Sestito2021}.
Comparing the properties of these galaxies with observations of the MW and local disk galaxies from the SPARC data \citep{Lelli2016}, \citet{Buck2020a} showed that simulated galaxy properties agree well with observations.

\begin{figure*}
    \script{half_mass_radius_metal_gradient.py}
    \begin{centering}
     \includegraphics[width=\linewidth]{figures/half_mass_radius_metal_gradient.pdf}
        \vspace*{-1.75em}
        \caption{
            Half mass radius (orange line, right axis) and metallicity gradient (blue line, left axis) of cold ($T<$~15000~K) gas as a function of time. The sharp increases in the cold gas disc size around the times of gas rich mergers coincide with a sudden steepening of the metallicity gradient of $\sim0.02$ dex/kpc. The addition of relatively unenriched cold gas by the merging satellite leads to a sudden increase of the star-forming cold gas mass in the outskirts of the galaxies while the central parts are unaffected. This steepens the [Fe/H] gradient. Gray shaded areas are the same as in Fig.~\ref{fig:merger_ratio} and mark the times at which early massive and gas rich mergers impact strongly the cold gas disk size.
        }
        \label{fig:half_mass}
    \end{centering}
\end{figure*}

The simulations assume cosmological parameters from the \cite{Planck}, namely: \OmegaM= 0.3175, \OmegaL= 0.6825, \Omegab= 0.049, H${_0}$ = 67.1\kms\Mpc$^{-1}$, \sig8 = 0.8344. Initial conditions are created the same way as for the original NIHAO runs \citep[see][]{Wang2015} using a modified version of the \texttt{GRAFIC2} package \citep{Bertschinger2001,Penzo2014}. The mass resolution of these simulations ranges between $m_{\rm dark}\sim1.5 - 5.1\times10^5 \Msun$ for dark matter particles and $m_{\rm gas}\sim2.8 - 9.4\times10^4 \Msun$ for the gas particles. The corresponding force softenings are $\epsilon_{\rm dark}=414 - 620$ pc for the dark matter and $\epsilon_{\rm gas}=177 - 265$ pc for the gas and star particles. However, the adaptive smoothing length scheme implies that $h_{\rm  smooth}$ can be as small as $\sim30$ pc in the disk mid-plane. Stellar particles are born with an initial mass of $1/3\times m_{\rm{gas}}$ and are subject to massloss according to stellar evolution models as detailed in \citet{Stinson2013}. 
The simulation setup, star formation and feedback implementations are described in detail in the introductory paper \citep{Buck2020a} but for completeness we summarise them below.

@@

Following \citet{Stinson2013} two modes of stellar feedback are implemented: (i) the energy input from young massive stars, e.g. stellar winds and photo ionization, prior to any supernovae explosions, thus termed \textit{early stellar feedback} (ESF). This mode consists of the total stellar luminosity ($2 \times 10^{50}$ erg of thermal energy per $M_{\odot}$) of the entire stellar population with an efficiency for coupling the energy input of $\epsilon_{\rm ESF}=13\%$ \citep{Wang2015}. (ii) supernova explosions implemented using the blastwave formalism as described in \citet{Stinson2006} and making use of a delayed cooling formalism for particles inside the blast region following \citet{McKee1977} in order to account for the adiabatic expansion of the supernova.
Finally, we adopted a metal diffusion algorithm between particles as described in \citet{Wadsley2008}.

The haloes in the zoom-in simulations were identified using the MPI+OpenMP hybrid halo finder \texttt{AHF2} \citep{Knollmann2009} and we use the accompanying merger tree tool to trace the particle IDs of all dark matter particles through time and identify all progenitor halos of a given galaxy/dark matter halo at redshift $z=0$. Subsequent analysis of the merger tree files is then performed with the {\texttt {ytree}} package \citep{ytree}.

\begin{figure}
    \script{gas_profile_all.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/gas_profile_2d.pdf}
        \vspace*{-1.75em}
        \caption{
            Relative change in cold gas surface density profile before and after the merger events as indicated in Fig.~\ref{fig:merger_ratio} by the gray vertical lines. Outside of a radius of $\sim5-10$ kpc the merging satellites contribute significant amounts of cold gas up to 200\%.
        }
        \label{fig:surf_den}
    \end{centering}
\end{figure}

@@

\subsection{The merger history} \label{sec:mergerhist}

We start our analysis on the influence of Gaia-Sausage-Enceladus-like merger
events on the chemo-dynamics of MW-like galaxy disks by looking at the merger history of the four galaxies. Figure~\ref{fig:merger_ratio} shows the cumulative gas mass growth (blue line, left axis) accompanied by the gaseous merger ratio (orange line, right axis) defined as the ratio of gas mass of the merging satellite over the gas mass of the main galaxy measured at the time of the merger. We see that after an initial phase ($0-\sim2.5$ Gyr) of rapid gas mass growth accompanied by violent, relatively gas rich mergers there are several other mergers at later times that contribute more than 10\% in gas (as measured at the time of merging) leading to sudden jumps in the gas mass of the main galaxy (blue line). 
Especially for g2.79e12 (left most panel) and g7.08e11 (right most panel) we see some late time ($\sim10$ Gyr) gas rich merger while g8.26e11 and g7.55e11 do not show any gas rich merger after $5$ and $7$ Gyr, respectively. For the purpose of this paper, we highlight important gas rich merger events i.e. those that lead to a strong increase in the cold gas disc size (see also Fig.~\ref{fig:half_mass}) by gray shaded areas in Figs.~\ref{fig:merger_ratio}, \ref{fig:half_mass} and \ref{fig:feh_evolution}. Note that there might be more gas rich mergers especially at early times but our focus here is on the particular mergers that heavily impact the cold gas disk size.

\begin{figure*}
    \script{enrichment_evolution.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/enrichment_evolution.pdf}
        \vspace*{-1.75em}
        \caption{
            Cold gas metallicity difference $\Delta\mathrm{[Fe/H]}=\mathrm{[Fe/H]}(R)-\mathrm{[Fe/H]}(R<2\, \mathrm{kpc})$, between the gas in a given annulus at a radius $R$ (colored lines) and the metallicity in the central parts ($R<2$ kpc) of the galaxy, $\mathrm{[Fe/H]}(R<2\, \mathrm{kpc})$ as a function of time. A flat line indicates perfect self-similar evolution between two different radii while a more negative value indicates that a certain radius lacks behind the central galaxy in [Fe/H] enrichment and is thus indicative of a steepening of the metallicity gradient. Gray shaded areas mark times where the gas disk size sharply increases due to gas rich mergers (see Fig.~\ref{fig:half_mass}).
        }
        \label{fig:feh_evolution}
    \end{centering}
\end{figure*}
